---
title: "Population Uncertainty Project"
author: "Armelle Duston"
date: "`r Sys.Date()`"
output: pdf_document
---

## Setup and Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(warning = F, message = F, echo = F)

### Load Packages 
library(ggplot2)
library(tidyr)
library(patchwork)
library(dplyr)
library(knitr)
library(kableExtra)
library(MASS)
library(nimble)
library(coda)
library(extraDistr)
library(igraph)
library(RColorBrewer)
library(nimbleNoBounds)
library(pracma)
library(extraDistr)

### ggplot theme
theme_set(theme_bw() +
            theme(plot.title = element_text(hjust = 0.5, size = 16),
                  plot.subtitle = element_text(hjust = 0.5, size = 12),
                  axis.title = element_text(size = 12),
                  axis.text = element_text(size = 10),
                  strip.text = element_text(size = 12),
                  legend.position = "bottom"))

if (rstudioapi::isAvailable()) {
  setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
}
```


## Generate Data

```{r generate_data}

source("src/generate_data.r")

n <- 100 # number of points
rho <- 0.7 # spatial correlation       
kappa <- 1.5 # scalar precision of the Gaussian CAR prior
tau <- 0.25 # related to census privacy budget     
J <- 10  # number of regions


sim_data <- generate_data(n=n, 
                          rho=rho, 
                          kappa=kappa, 
                          tau=tau,
                          J=J)


```

### Plot neighborhood structure

```{r plot_neighborhood}

# sim_data$W       : n×n adjacency matrix
# sim_data$S       : latent log-intensities
# sim_data$P       : true counts
# sim_data$region_id: region labels  


# Build igraph object
g <- graph_from_adjacency_matrix(sim_data$W, mode = "undirected")

# Layout (e.g. Fruchterman–Reingold)
layout_ij <- layout_with_fr(g)

# 1. Choose a palette with J distinct colors
J <- length(unique(sim_data$region_id))
palette <- brewer.pal(min(J, 12), "Set3")  # up to 12 colors; you can pick another brewer palette

# 2. Assign each region its color
region_cols <- palette[sim_data$region_id]

# 3. Plot, coloring nodes by region
plot(g, 
     layout        = layout_ij,
     vertex.size   = 6,
     vertex.color  = region_cols,
     vertex.label  = NA,
     edge.color    = "grey80",
     main          = "Adjacency Graph with Regions")

# 4. Add a legend
legend("topright",
       legend = paste("Region", sort(unique(sim_data$region_id))),
       col    = palette[sort(unique(sim_data$region_id))],
       pch    = 19,
       pt.cex = 1.5,
       bty    = "n")


```



## Spatial-only Bayesian Model

Get just the spatial piece working

```{r spatial_only}

source("src/spatial_only.r")

samples1 <- spatial_only(sim_data)

# Diagnostics
traceplot(as.mcmc.list(samples1)[, c("rho", "log_kappa", "S[1]", "S[25]", "S[50]")])

gelman.diag(as.mcmc.list(samples1)[, c("rho", "log_kappa", "S[1]", "S[25]", "S[50]")], 
            autoburnin = FALSE)
effectiveSize(as.mcmc.list(samples1)[, c("rho", "log_kappa", "S[1]", "S[25]", "S[50]")])
  
  
```



## Spatial + DP noise + benchmarking Bayesian Model


```{r ddlaplace_nimble_function}

# Define the ddlaplace_nim distribution for NIMBLE
ddlaplace_nim <- nimbleFunction(
  run = function(x = double(0), 
                 mu = double(0), 
                 tau = double(0), 
                 log = integer(0, default = 0)) {
    returnType(double(0))
    
    # Parameter checks
    if(tau <= 0) return(NaN)
    
    # Calculate lambda parameter (lambda = exp(-1/tau))
    lambda <- exp(-1/tau)
    
    # Calculate normalization constant
    c <- (1 - lambda) / (1 + lambda)
    
    # Calculate PMF
    logProb <- log(c) + abs(x - mu) * log(lambda)
    
    # Return log or natural probability
    if(log) return(logProb)
    else return(exp(logProb))
  }
)

# Define the simulation function without using rgeom
rdlaplace_nim <- nimbleFunction(
  run = function(n = integer(0), mu = double(0), tau = double(0)) {
    returnType(double(0))
    
    # Parameter check
    if(tau <= 0) return(NaN)
    if(n != 1) stop("rdlaplace_nim only handles n = 1")
    
    # Calculate lambda
    lambda <- exp(-1/tau)
    
    # Probability of X <= mu
    p_leq_mu <- 1 / (1 + lambda)
    
    # Decide direction using uniform random number
    u <- runif(1)
    if(u < p_leq_mu) {
      # X <= mu (including equality)
      # Instead of using rgeom, implement geometric sampling directly
      v <- runif(1)
      # Convert uniform to geometric: k = floor(log(v) / log(lambda))
      k <- floor(log(v) / log(1 - lambda))
      return(mu - k)
    } else {
      # X > mu
      v <- runif(1)
      # Convert uniform to geometric: k = floor(log(v) / log(lambda))
      k <- floor(log(v) / log(1 - lambda))
      return(mu + k + 1)
    }
  }
)

# Register the distribution with NIMBLE
registerDistributions(list(
  ddlaplace_nim = list(
    BUGSdist = "ddlaplace_nim(mu, tau)",
    discrete = TRUE,
    types = c('value = double(0)', 'mu = double(0)', 'tau = double(0)'),
    pqAvail = FALSE,
    range = c(-Inf, Inf)
  )
))
```

### Complete model with exact benchmarking

```{r exact_bench}
source("src/complete_model.r")

samples2 <- complete_model(sim_data, exact = TRUE)

# Diagnostics
traceplot(as.mcmc.list(samples2)[, c("rho", "log_kappa", 
                                     "S[1]", "S[5]", "S[50]", 
                                     "P[1]", "P[5]", "P[50]")])

gelman.diag(as.mcmc.list(samples2)[, c("rho", "log_kappa", 
                                     "S[1]", "S[5]", "S[50]", 
                                     "P[1]", "P[5]", "P[50]")], 
            autoburnin = FALSE)
effectiveSize(as.mcmc.list(samples2)[, c("rho", "log_kappa", 
                                     "S[1]", "S[5]", "S[50]", 
                                     "P[1]", "P[5]", "P[50]")])

```

### Complete model with inexact benchmarking

```{r}

samples3 <- complete_model(sim_data, exact = FALSE)

# Diagnostics
traceplot(as.mcmc.list(samples3)[, c("rho", "log_kappa", 
                                     "S[1]", "S[5]", "S[50]", 
                                     "P[1]", "P[5]", "P[50]")])

gelman.diag(as.mcmc.list(samples3)[, c("rho", "log_kappa", 
                                     "S[1]", "S[5]", "S[50]", 
                                     "P[1]", "P[5]", "P[50]")], 
            autoburnin = FALSE)
effectiveSize(as.mcmc.list(samples3)[, c("rho", "log_kappa", 
                                     "S[1]", "S[5]", "S[50]", 
                                     "P[1]", "P[5]", "P[50]")])


```


## Model Evaluation 

Compare modeled vs true P's

```{r}

true_Ps <- sim_data$P
obs_Ps <- sim_data$P_star
modeled_Ps_exact <- samples2$chain1[,paste0("P[", 1:100, "]")]
modeled_Ps_inexact <- samples3$chain1[,paste0("P[", 1:100, "]")]

# posterior median = observed? (if yes rmse's equal)
rmserr(true_Ps, obs_Ps)$rmse
rmserr(true_Ps, apply(X=modeled_Ps_exact, MARGIN=2 , FUN=median))$rmse
rmserr(true_Ps, apply(X=modeled_Ps_inexact, MARGIN=2 , FUN=median))$rmse

# 95% CI coverage
CIs_exact <- apply(X=modeled_Ps_exact, MARGIN=2 , 
      FUN= function(x) quantile(x, probs = c(0.025, 0.975))) 

print("Exact benchmarking coverage")
sum(ifelse(true_Ps >= CIs_exact[1,] & true_Ps <= CIs_exact[2,], 1, 0))/100

CIs_inexact <- apply(X=modeled_Ps_inexact, MARGIN=2 , 
      FUN= function(x) quantile(x, probs = c(0.025, 0.975))) 

print("Inexact benchmarking coverage")
sum(ifelse(true_Ps >= CIs_inexact[1,] & true_Ps <= CIs_inexact[2,], 1, 0))/100


```

#### INEXACT BENCHMARKING NOT WORKING ######


## Questions
- Appropriate to sample using discrete laplace? Or better to use laplace and then round? 
- Does the constraint approach make sense? 
